---
title: "Survey and experiment pilots with LLMRpanel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey and experiment pilots with LLMRpanel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Set RUN_LIVE <- TRUE to evaluate chunks with configured provider credentials.
# FALSE keeps package and site builds offline.
RUN_LIVE <- FALSE
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = RUN_LIVE)
```

LLMRpanel administers survey and experimental instruments to panels of
language model personas. Use it to pretest questions, pilot conjoint
designs, calculate sample sizes from pilot dispersion, or measure responses
from a configured model. `panel_benchmark()` compares closed-item response
shares with benchmark shares supplied by the user. It records deviations by
item and response, nonresponse, and the number of closed items covered.

Set `RUN_LIVE` to `TRUE` to evaluate the provider calls below. The default
`FALSE` setting lets package and site builds run without provider credentials
or charges.

## What silicon panels are for

1. **Instrument pretesting.** Administer draft items and inspect unmatched
replies and first-option sensitivity.

2. **Design piloting.** Run conjoint tasks and estimate response
dispersion before planning human data collection.

3. **Model measurement.** Compare response distributions across personas,
item orders, option orders, or model configurations.

## Panels and instruments

```{r setup}
library(LLMRpanel)
```

`panel_from_margins()` samples attribute values from the supplied marginal
distributions. `set.seed()` makes this draw reproducible.

```{r panel, purl=FALSE}
set.seed(110)
panel = panel_from_margins(
  list(
    age = c("18 to 34" = .30, "35 to 64" = .45, "65 plus" = .25),
    party = c(left = .45, right = .45, independent = .10)
  ),
  n = 12,
  persona_template = "A {age} year old voter who leans {party}."
)
panel

instrument = panel_instrument(list(
  item_likert("wk4", "A four day work week would benefit society."),
  item_choice(
    "fund",
    "Which should the city fund first?",
    c("public transit", "road repair")
  ),
  item_open("why", "In one sentence, why?")
))
instrument
```

Margins are useful when targets are published as tables. When microdata is
available, `panel_from_data()` is the joint distribution counterpart. It
draws personas from observed rows and therefore preserves relationships
among attributes rather than sampling each margin independently.
`LLMR::report()` identifies whether a panel came from supplied margins,
microdata rows, or supplied personas.

`panel_administer()` sends each item to each persona. It returns a
`panel_responses` object with response rows in `$data` and the panel,
instrument, benchmark record, and token usage in ordinary fields. It randomizes
item and option order by default and records `item_position` and `option_order`
in `$data`. The example repeats the administration with a second model
configuration so response patterns can be compared across models.

```{r administer, purl=FALSE}
cfg = LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.8)
cfg_qwen = LLMR::llm_config("groq", "qwen/qwen3-32b", temperature = 0.8)

resp = panel_administer(panel, instrument, cfg)
resp
resp$data
panel_bias_audit(resp)
LLMR::diagnostics(resp)

resp_qwen = panel_administer(panel, instrument, cfg_qwen)
panel_bias_audit(resp_qwen)
```

## Compare responses with a benchmark

`panel_benchmark()` compares valid model response shares with benchmark
shares for matching item-response pairs. It also records benchmark coverage
and item-level nonresponse in `$benchmark`. Before a benchmark is attached,
response shares describe the configured model under the supplied personas,
not a human population. `bench_fund` supplies shares for one closed item.

```{r benchmark_partial, purl=FALSE}
bench_fund = data.frame(
  item_id = rep("fund", 2),
  response = c("public transit", "road repair"),
  share = c(0.41, 0.59)
)

resp_partial = panel_benchmark(
  resp,
  bench_fund,
  benchmark_name = "toy city survey"
)
resp_partial
```

`bench_fund` covers one of the instrument's two closed items. `bench_all`
adds shares for `wk4` and covers both.

```{r benchmark_full, purl=FALSE}
bench_all = rbind(
  bench_fund,
  data.frame(
    item_id = rep("wk4", 5),
    response = c(
      "strongly disagree",
      "disagree",
      "neutral",
      "agree",
      "strongly agree"
    ),
    share = c(.05, .20, .25, .35, .15)
  )
)

resp = panel_benchmark(
  resp,
  bench_all,
  benchmark_name = "toy city survey"
)
resp
LLMR::report(resp)
resp$benchmark$nonresponse
```

`resp$benchmark$nonresponse` gives the missing response proportion for each
closed item. The comparison shares use nonmissing responses as their
denominator.

## Conjoint designs

`conjoint_design()` uses R's random-number generator to construct a classed
design list. Its `$profiles` field contains the initial profile table, and its
`$attributes` field contains the attribute universe. `panel_administer()` draws
the profiles each respondent sees. Set a seed before administration to
reproduce those respondent-level draws.

```{r conjoint_design, purl=FALSE}
set.seed(110)
design = conjoint_design(
  list(
    price = c("low", "high"),
    origin = c("domestic", "imported")
  ),
  n_tasks = 4
)
design
design$profiles
design$attributes
```

`conjoint_design()` attempts to use distinct profiles within each task in
`$profiles` and warns when the attribute space cannot supply them.
`conjoint_instrument()` creates one forced-choice item per task. Administration
renders a fresh draw for each respondent and records it with the response.
`conjoint_amce()` estimates from those recorded profiles relative to the first
level of each attribute and calculates standard errors clustered by persona.

```{r conjoint_amce, purl=FALSE}
cj_instr = conjoint_instrument(design, "Which product would you buy?")
cj = panel_administer(panel, cj_instr, cfg)
conjoint_amce(cj)
```

`conjoint_amce()` returns a classed result with one row for each observed
attribute level. Baseline levels have estimate 0 and missing standard errors.
Other rows contain the estimated contrast and 95 percent interval. Run counts
remain in ordinary columns.

## Two-arm sample sizes

`panel_power()` computes two-arm sample sizes from a minimum detectable effect
and pilot dispersion. It uses the standard deviation of Likert scores. For
choice items, it uses a named focal response or the modal response share. The
result has one row per analyzed closed item.

```{r power, purl=FALSE}
panel_power(resp, effect = c(wk4 = 0.4, fund = 0.15))
```

## Request counts and model choice

`panel_administer()` makes one request per persona-item pair. Item and option
randomization do not add requests. Provider prices and prompt and response
lengths determine cost. A versioned local model can support later reruns when
hosted endpoints change. The `$data` field retains `response_text`,
`response_id`, `success`, `model`, and `provider`, including when a reply cannot
be matched to a closed-item option. `finish_reason` is retained when the runner
supplies it.

## Relations

[LLMR](https://asanaei.github.io/LLMR/) supplies provider configuration and
execution. [LLMRcontent](https://asanaei.github.io/LLMRcontent/) provides
codebook-based text annotation and validation.
[LLMRagent](https://asanaei.github.io/LLMRagent/) provides agent experiments.
[LLMRpanel](https://asanaei.github.io/LLMRpanel/) contains panel constructors,
instruments, administration, and response summaries.
