---
title: "LLMRpanel by design: silicon samples that know their place"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLMRpanel by design: silicon samples that know their place}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = identical(tolower(Sys.getenv("LLMRPANEL_RUN_VIGNETTES", "false")), "true")
)
```

This vignette is the package's design document. The literature on silicon
sampling is genuinely split: persona-conditioned models reproduce some
human marginals strikingly well and fail others silently -- over-uniform
variances, sanitized refusals, order effects, sensitivity to persona
phrasing. Both findings are real. The design conclusion is not "never
simulate" but "never simulate *uncalibrated*" -- and the package encodes
that conclusion in its objects rather than in its documentation: every
result prints an **UNCALIBRATED** banner until a benchmark comparison
fills the calibration slot.

## What silicon panels are for

The high-payoff uses do not require believing the model is a population:

- **Instrument pretesting.** Confusing items, broken branching, attribute
  levels nobody distinguishes -- caught overnight, for the price of
  electricity, before a single human respondent's time is spent.
- **Design piloting.** Run 40 variants of a conjoint or vignette design;
  estimate design effects for the power analysis; file the silicon pilot
  in the preregistration.
- **Model measurement.** How a model answers attitude items under varied
  personas is a finding *about the model* -- bias audits as research.

Estimating human quantities is the use that requires earning, and
`calibrate()` is where the earning is measured, item by item, deviation
by deviation.

## A walk through the machinery

```{r setup}
library(LLMRpanel)
set.seed(110)   # panels and order randomization draw locally

panel <- panel_from_margins(
  list(age   = c("18-34" = .30, "35-64" = .45, "65+" = .25),
       party = c(left = .45, right = .45, independent = .10)),
  n = 60,
  persona_template = "A {age}-year-old voter who leans {party}."
)
panel

instr <- instrument(list(
  item_likert("wk4", "A four-day work week would benefit society."),
  item_choice("fund", "Which should the city fund first?",
              c("public transit", "road repair")),
  item_open("why", "In one sentence, why?")
))
instr
```

Administration randomizes item and option order per respondent and
records what each one saw -- with LLMs, option order is a treatment:

```{r administer}
cfg  <- LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.8)
resp <- administer(panel, instr, cfg)
resp          # prints the UNCALIBRATED banner, by design
bias_audit(resp)
```

Calibration is comparison, not adjustment -- deviations are reported,
never massaged away:

```{r calibrate}
bench <- data.frame(
  item_id  = c("fund", "fund"),
  response = c("public transit", "road repair"),
  share    = c(0.41, 0.59)          # from your human benchmark source
)
resp <- calibrate(resp, bench, benchmark_name = "city survey 2025")
resp          # banner replaced by the measured deviation
panel_report(resp)
```

## Factorial stimuli

```{r designs}
vignette_design(
  "A {age} applicant with {experience} of experience applies.",
  list(age = c("younger", "older"),
       experience = c("5 years", "20 years"))
)

conjoint_design(
  list(price = c("low", "high"), origin = c("domestic", "imported")),
  n_tasks = 3
)
```

Pair either with an `item_choice()` and administer; AMCE estimation
(`amce()`, delegating to the established conjoint stack) arrives in 0.2.

## Why open weights matter here specifically

Panels are call-hungry: personas x items x orders x replicates. Three
things change at open-weight prices. Volume stops being the constraint --
a 2,000-persona pretest overnight costs electricity. `mode = "logprob"`
(0.2) reads the full response-option distribution from one forward pass --
reliably available only where logprobs are exposed, which in practice
means open-weight backends. And a panel pinned to an open checkpoint is a
*permanent instrument*: the same personas can answer the same items in
ten years, which no deprecated API model can promise.

## Arriving in 0.2 (contracts already exported)

- `administer(mode = "logprob")` -- option distributions without sampling.
- `persona_paraphrase()` -- persona wording as a sensitivity arm.
- `silicon_power()` -- power analysis for the *human* study, with
  silicon-informed priors; the clearest payoff per dollar.
- `amce()` -- conjoint estimation via cregg/marginaleffects.
- Joint persona draws from microdata, and raking to known cross-tabs.

## Relations

FocusGroup (same author) owns qualitative group simulation; LLMRpanel is
strictly quantitative. Persona phrasing and option order are exactly the
perturbations `LLMRvalid` audits; administrations land in the audit log
and therefore in `LLMRtrail` archives; open-ended answers get coded with
`LLMRcoder` -- under a codebook, against gold, like any other text.
