---
title: "LLMRpanel by design: calibrated silicon samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLMRpanel by design: calibrated silicon samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Set RUN_LIVE <- TRUE once, with API keys available, to execute every chunk
# against real models and confirm the vignette's outputs are honest. It is
# committed as FALSE so routine package and site builds do not require keys
# or spend; the prose describes what the live run shows.
RUN_LIVE <- FALSE
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = RUN_LIVE)
```

This vignette is the package design document. The literature on silicon
sampling is genuinely split. Persona conditioned models reproduce some
human marginals strikingly well and fail others silently through over
uniform variances, sanitized refusals, order effects, and sensitivity to
persona phrasing. Both findings are real. The design conclusion is not
"never simulate" but "never simulate uncalibrated", and the package encodes
that conclusion in its objects rather than in a warning paragraph. Every
result prints an **UNCALIBRATED** banner until a benchmark comparison fills
the calibration slot, and partial benchmarks earn only a **PARTIALLY
CALIBRATED** one.

The chunks below are live model calls when `RUN_LIVE` is set to `TRUE`.
They use real `LLMR::llm_config()` objects on the LLMR provider layer and
call execution functions without a runner argument. With the switch left
as `FALSE`, the document builds without keys or model spend, while the
prose explains what the live run is meant to show.

## What silicon panels are for

1. **Instrument pretesting.** Confusing items, broken branching, and
attribute levels nobody distinguishes can be caught before any human
respondent's time is spent.

2. **Design piloting.** A conjoint or vignette design can be exercised
across many personas so the analyst can inspect response dispersion and
price the later human study.

3. **Model measurement.** How a model answers attitude items under varied
personas is a finding about the model. Bias audits are research objects,
not merely quality checks.

Estimating human quantities is the use that requires earning.
`panel_calibrate()` is where the earning is measured, item by item and deviation
by deviation.

## A walk through the machinery

```{r setup}
library(LLMRpanel)
```

`panel_from_margins()` draws a silicon panel from marginal targets. The
seed is set here because the panel draw is local random generation that
the vignette should reproduce.

```{r panel}
set.seed(110)
panel = panel_from_margins(
  list(
    age = c("18 to 34" = .30, "35 to 64" = .45, "65 plus" = .25),
    party = c(left = .45, right = .45, independent = .10)
  ),
  n = 12,
  persona_template = "A {age} year old voter who leans {party}."
)
panel

instr = panel_instrument(list(
  item_likert("wk4", "A four day work week would benefit society."),
  item_choice(
    "fund",
    "Which should the city fund first?",
    c("public transit", "road repair")
  ),
  item_open("why", "In one sentence, why?")
))
instr
```

Margins are useful when targets are published as tables. When microdata is
available, `panel_from_data()` is the joint distribution counterpart. It
draws personas from observed rows and therefore preserves relationships
among attributes rather than sampling each margin independently.

Administration randomizes item and option order per respondent and records
what each one saw. With language models, option order is a treatment. The
first configuration is the main pilot model. The second family is useful
when the question is whether a pattern is tied to one model family or is
visible across families.

```{r administer}
cfg = LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.8)
cfg_qwen = LLMR::llm_config("groq", "qwen/qwen3-32b", temperature = 0.8)

resp = panel_administer(panel, instr, cfg)
resp
panel_bias_audit(resp)
LLMR::diagnostics(resp)

resp_qwen = panel_administer(panel, instr, cfg_qwen)
panel_bias_audit(resp_qwen)
```

The first printed response object reports **UNCALIBRATED**. That is the
intended default. The comparison model is not a human benchmark; it is a
model measurement contrast.

## Calibration is comparison, not adjustment

Deviations are reported, not massaged away. The comparison is restricted
to items the benchmark actually covers, and coverage is part of the
verdict. The following toy benchmark is inline so the example is complete.
It demonstrates the mechanics of calibration but is not evidence about a
real population.

```{r calibrate_partial}
bench_fund = data.frame(
  item_id = rep("fund", 2),
  response = c("public transit", "road repair"),
  share = c(0.41, 0.59)
)

resp_partial = panel_calibrate(
  resp,
  bench_fund,
  benchmark_name = "toy city survey"
)
resp_partial
```

Because this benchmark touches only one of the closed items, the object
earns only **PARTIALLY CALIBRATED**. Covering every closed item lets the
object report the measured deviation for this toy benchmark in place of
the uncalibrated banner.

```{r calibrate_full}
bench_all = rbind(
  bench_fund,
  data.frame(
    item_id = rep("wk4", 5),
    response = c(
      "strongly disagree",
      "disagree",
      "neutral",
      "agree",
      "strongly agree"
    ),
    share = c(.05, .20, .25, .35, .15)
  )
)

resp = panel_calibrate(
  resp,
  bench_all,
  benchmark_name = "toy city survey"
)
resp
panel_report(resp)
LLMR::report(resp)
attr(resp, "calibration")$nonresponse
```

Nonresponse is recorded per item. Shares computed only over valid answers
can flatter an instrument the model often refuses, so the calibration
object keeps the missingness visible.

## Factorial stimuli

Factorial designs are local design objects. A vignette design expands a
template over attribute levels.

```{r designs}
vignette_design(
  "A {age} applicant with {experience} of experience applies.",
  list(
    age = c("younger", "older"),
    experience = c("5 years", "20 years")
  )
)
```

A conjoint design draws randomized tasks. The seed is set here because the
design draw is local random generation.

```{r conjoint_design}
set.seed(110)
design = conjoint_design(
  list(
    price = c("low", "high"),
    origin = c("domestic", "imported")
  ),
  n_tasks = 4
)
design
```

Profiles within a conjoint task are guaranteed distinct, with a warning
when the attribute space is too small to allow it. `conjoint_instrument()`
renders the design into one forced choice item per task. After live
administration, `amce()` estimates average marginal component effects with
respondent clustered standard errors.

```{r amce}
cj_instr = conjoint_instrument(design, "Which product would you buy?")
cj = panel_administer(panel, cj_instr, cfg)
amce(cj)
```

The AMCE table is a diagnostic of how this configured model responded to
the randomized profiles. A price effect appears only if the live choices
express such a preference; an origin effect appears only if the model
distinguishes that attribute.

## Pricing the human study

A silicon pilot prices the design before any human is recruited.
`panel_power()` reads dispersion from the pilot, such as the standard
deviation of a Likert score or the modal share of a choice, and returns
the per arm sample size a human study would need to detect a specified
effect. The number inherits the pilot's calibration status. With the toy
benchmark above, it is a design calculation rather than external
validation.

```{r power}
panel_power(resp, effect = c(wk4 = 0.4, fund = 0.15))
```

## Why open weights matter here specifically

Panels are call hungry: personas by items by orders by replicates. Open
weight prices change what can be piloted. Volume stops being the primary
constraint, and a panel pinned to an open checkpoint becomes a durable
instrument. The same personas can answer the same items later, which no
deprecated API model can promise.

## Relations

[LLMRpanel](https://asanaei.github.io/LLMRpanel/) is the quantitative
member of the family. [LLMR](https://asanaei.github.io/LLMR/) supplies the
provider abstraction used by the live calls above. Persona phrasing and
option order are the same treatments
[LLMRvalid](https://asanaei.github.io/LLMRvalid/) audits. Administrations
can land in an [LLMRarchive](https://asanaei.github.io/LLMRarchive/) archive.
Open answers can be coded in
[LLMRcoder](https://asanaei.github.io/LLMRcoder/) under a codebook and
against gold text. [LLMRAgent](https://asanaei.github.io/LLMRAgent/)
addresses agent style experiments, while this package keeps attention on
survey and experiment samples.
