---
title: "LLMRpanel by design: silicon samples that know their place"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLMRpanel by design: silicon samples that know their place}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette is the package's design document. The literature on silicon
sampling is genuinely split: persona-conditioned models reproduce some
human marginals strikingly well and fail others silently -- over-uniform
variances, sanitized refusals, order effects, sensitivity to persona
phrasing. Both findings are real. The design conclusion is not "never
simulate" but "never simulate *uncalibrated*" -- and the package encodes
that conclusion in its objects rather than in its documentation: every
result prints an **UNCALIBRATED** banner until a benchmark comparison
fills the calibration slot, and partial benchmarks earn only a
**PARTIALLY CALIBRATED** one.

Everything below **runs offline**: a simulated respondent (a deterministic
function on the `.runner` seam) stands in for the model, which also makes
the document a worked example of testing panel pipelines without spending
a token. The final section swaps in a real open-weight model.

```{r setup}
library(LLMRpanel)
set.seed(110)   # panels and order randomization draw locally

simulated_respondent <- function(experiments, ...) {
  experiments$response_text <- vapply(seq_len(nrow(experiments)), function(i) {
    sys <- experiments$messages[[i]][["system"]]
    usr <- experiments$messages[[i]][["user"]]
    right <- grepl("leans right", sys)
    if (grepl("work week", usr))   { if (right) "agree" else "disagree" }
    else if (grepl("fund first", usr)) { if (right) "road repair" else "public transit" }
    else "Because it fits how I live."
  }, character(1))
  experiments
}
```

## What silicon panels are for

The high-payoff uses do not require believing the model is a population:

- **Instrument pretesting.** Confusing items, broken branching, attribute
  levels nobody distinguishes -- caught overnight, for the price of
  electricity, before a single human respondent's time is spent.
- **Design piloting.** Run 40 variants of a conjoint or vignette design;
  estimate design effects for the power analysis; file the silicon pilot
  in the preregistration.
- **Model measurement.** How a model answers attitude items under varied
  personas is a finding *about the model* -- bias audits as research.

Estimating human quantities is the use that requires earning, and
`calibrate()` is where the earning is measured, item by item, deviation
by deviation.

## A walk through the machinery

```{r panel}
panel <- panel_from_margins(
  list(age   = c("18-34" = .30, "35-64" = .45, "65+" = .25),
       party = c(left = .45, right = .45, independent = .10)),
  n = 40,
  persona_template = "A {age}-year-old voter who leans {party}."
)
panel

instr <- instrument(list(
  item_likert("wk4", "A four-day work week would benefit society."),
  item_choice("fund", "Which should the city fund first?",
              c("public transit", "road repair")),
  item_open("why", "In one sentence, why?")
))
instr
```

Administration randomizes item and option order per respondent and
records what each one saw -- with LLMs, option order is a treatment:

```{r administer}
cfg  <- LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.8)
resp <- administer(panel, instr, cfg, .runner = simulated_respondent)
resp          # prints the UNCALIBRATED banner, by design
bias_audit(resp)
```

## Calibration is comparison, not adjustment

Deviations are reported, never massaged away; the comparison is
restricted to items the benchmark actually covers; and coverage is part
of the verdict. A benchmark touching one of two closed items earns
**PARTIALLY CALIBRATED**, not a clean bill:

```{r calibrate-partial}
bench_fund <- data.frame(
  item_id  = "fund",
  response = c("public transit", "road repair"),
  share    = c(0.41, 0.59)          # from your human benchmark source
)
resp <- calibrate(resp, bench_fund, benchmark_name = "city survey 2025")
resp
```

Cover everything closed and the banner retires, the measured deviation
standing in its place:

```{r calibrate-full}
bench_all <- rbind(bench_fund, data.frame(
  item_id  = "wk4",
  response = c("strongly disagree", "disagree", "neutral",
               "agree", "strongly agree"),
  share    = c(.05, .20, .25, .35, .15)))
resp <- calibrate(resp, bench_all, benchmark_name = "city survey 2025")
resp
panel_report(resp)
```

Nonresponse is recorded per item alongside
(`attr(resp, "calibration")$nonresponse`): shares computed only over
valid answers flatter an instrument the model often refuses.

## Factorial stimuli

```{r designs}
vignette_design(
  "A {age} applicant with {experience} of experience applies.",
  list(age = c("younger", "older"),
       experience = c("5 years", "20 years"))
)

conjoint_design(
  list(price = c("low", "high"), origin = c("domestic", "imported")),
  n_tasks = 3
)
```

Profiles within a conjoint task are guaranteed distinct -- a forced choice
between clones measures nothing -- with a warning when the attribute space
is too small to allow it. Pair either design with an `item_choice()` and
administer; AMCE estimation (`amce()`, delegating to the established
conjoint stack) arrives in 0.2.

## Why open weights matter here specifically

Panels are call-hungry: personas x items x orders x replicates. Three
things change at open-weight prices. Volume stops being the constraint --
a 2,000-persona pretest overnight costs electricity. A logprob
administration path (0.2) will read the full response-option distribution
from one forward pass -- reliably available only where logprobs are
exposed, which in practice means open-weight backends. And a panel pinned
to an open checkpoint is a *permanent instrument*: the same personas can
answer the same items in ten years, which no deprecated API model can
promise.

## Live, when you are ready

Drop `.runner` and the same administration runs on the model in `cfg`.
With `GROQ_API_KEY` set and `LLMRPANEL_RUN_VIGNETTES=true`:

```{r live, eval = identical(tolower(Sys.getenv("LLMRPANEL_RUN_VIGNETTES", "false")), "true")}
administer(panel, instr, cfg)
```

## Arriving in 0.2 (contracts already exported)

- `silicon_power()` -- power analysis for the *human* study, with
  silicon-informed priors; the clearest payoff per dollar.
- `amce()` -- conjoint estimation via cregg/marginaleffects.
- A logprob administration path; persona-paraphrase sensitivity arms;
  joint persona draws from microdata, and raking to known cross-tabs.

## Relations

FocusGroup (same author) owns qualitative group simulation; LLMRpanel is
strictly quantitative. Persona phrasing and option order are exactly the
perturbations `LLMRvalid` audits; administrations land in the audit log
and therefore in `LLMRtrail` archives; open-ended answers get coded with
`LLMRcoder` -- under a codebook, against gold, like any other text.
